---
title: "Progress Memo 2"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: "Emily Lynott"
pagetitle: "PM2 Emily Lynott"
date: today

format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    embed-resources: true
    link-external-newwindow: true

execute:
  warning: false

from: markdown+emoji
reference-location: margin
citation-location: margin
---

::: {.callout-tip icon=false}

## Github Repo Link

[Emily's GitHub Repo](https://github.com/emilylynott/final-progress-memo-2.git)

:::

```{r}
#| label: loading
#| echo: false

library(tidyverse)

alcohol_gho <- readxl::read_xlsx("data/data.xlsx")

```
```{r}
#| label: cleaning
#| echo: false

alcohol_gho_tidy <- alcohol_gho |>
  pivot_wider(
    names_from = "dimension", 
    values_from = "subgroup") |>
  mutate(
    indicator_name = factor(indicator_name), 
    indicator_name = fct_recode(indicator_name, 
      "liters (pure alcohol) consumption per capita; 15+; drinkers only" = "Alcohol, drinkers only per capita (15+) consumption (in litres of pure alcohol)",
      "% abstainers, lifetime; NA; population" = "Alcohol, abstainers lifetime (%)",
      "% abstainers, past 12 mo; NA; population" = "Alcohol, abstainers past 12 months (%)",
      "% consumers, past 12 mo; NA; population" = "Alcohol, consumers past 12 months (%)",
      "% former drinkers; NA; population" = "Alcohol, former drinkers (%)", 
      "% heavy episodic drinkers, past 30 days; 15+; drinkers only" = "Alcohol, heavy episodic drinking (15+), drinkers only, past 30 days (%)", 
      "% dependence, 12 mo prevalence; 15+; population" = "Alcohol dependence (15+), 12 month prevalence (%)",
      "% use diorders, 12 mo prevalence; 15+; population" = "Alcohol use disorders (15+), 12 month prevalence (%)",
      "liters (pure alcohol) consumption per capita; 15+; population" = "Alcohol, total per capita (15+) consumption (SDG Indicator 3.5.2) (in litres of pure alcohol)", 
      "% heavy episodic drinkers, past 30 days; 15+; population, age standardized" = "Alcohol, heavy episodic drinking (15+) past 30 days (age-standardized) (%)", 
      "% alcohol-attributable fractions: liver cirrhosis deaths; 15+; population" = "Alcohol-attributable fractions (15+), liver cirrhosis deaths (%)",
      "% alcohol-attributable fractions: road traffic crash deaths; 15+; population" = "Alcohol-attributable fractions (15+), road traffic crash deaths (%)", 
      "% alcohol-attributable fractions: all-cause deaths; NA; population" = "Alcohol-attributable fractions, all-cause deaths (%)", 
      "grams average daily intake; NA; drinkers only" = "Alcohol, average daily intake among drinkers (grams)", 
      "% harmful use, 12 mo prevalence; 15+; population" = "Alcohol, harmful use (15+), 12 month prevalence (%)", 
      "% heavy episodic drinkers; 15-19; population" = "15-19 years old heavy episodic drinkers (population) (%)", 
      "% heavy episodic drinkers; 15-19; drinkers only" = "15-19 years old heavy episodic drinkers (drinkers only) (%)", 
      "% current drinkers; 15-19; population" = "15-19 years old, current drinkers (%)", 
      "% any alcoholic beverage consumed, past 12 mo; 15-19; population" = "15-years old any alcoholic beverage consumed in past 12 months (%)", 
      "% any alcoholic beverage consumed, past 30 days; 13-15; population" = "13-15-years old any alcoholic beverage consumed in past 30 days (%)", 
      "% first drink before 14; 13-15; population" = "13-15-years old first drink before age 14 (%)", 
      "% heavy episodic drinking, past 30 days; 15-19; population" = "Heavy episodic drinking (youth 15 -19 years) past 30 days (%)", 
      "% heavy episodic drinking, past 30 days; 15-19; drinkers only" = "Heavy episodic drinking (youth 15 -19 years), drinkers only, past 30 days (%)"
    ) 
  ) |>
  separate_wider_delim(
    indicator_name, 
    delim = ";", 
    names = c("indicator_behavior", "indicator_age", "indicator_population")
  ) |>
  mutate(
    indicator_behavior = factor(indicator_behavior)
  )
  
```

## General Population Analysis

I wanted to  begin my EDA looking at some of the variables that seemed to more generally describe each population. My main goal in doing this was to have a better understanding of how alcohol consumption is distributed, overall and in the context of some other factors. The initial distribution is analyzed through a boxplot and summary statistics below. 

```{r}
#| label: general
#| echo: false

tidy_general<- alcohol_gho_tidy |>
  filter(
    indicator_behavior %in% c(
      "% abstainers, lifetime", 
      "% abstainers, past 12 mo", 
      "% consumers, past 12 mo", 
      "% former drinkers"
    )
  )

tidy_general |>
  select(indicator_behavior, estimate) |>
  group_by(indicator_behavior) |>
  ggplot(aes(x = indicator_behavior, y = estimate)) +
  geom_boxplot() + 
  labs(
    title = "Distribution of Percents of Consumption Behaviors in Populations", 
    x = "Indicator Behavior", 
    y = "Estimated Percent")
  

tidy_general |>
  select(indicator_behavior, estimate) |>
  group_by(indicator_behavior) |>
  summarize(
    median_estimate = median(estimate, na.rm = TRUE), 
    IQR_estimate = IQR(estimate, na.rm = TRUE),
    min_estimate = min(estimate, na.rm = TRUE),
    max_estimate = max(estimate, na.rm = TRUE)
  ) |>
  knitr::kable(caption = "Table 1: Summary Statistics for percents of populations with alcohol consumption behaviors") 

```

This analysis displays that the median percent of consumers in the past 12 months is 37.5. This is very similar to the median percent of lifetime abstainers (39) but only about 60% of those abstaining for the past 12 months (62.35%). The median percent of former drinkers is just under half of the percent of consumers at 19.95%. 

## Income Analysis

The first variable I wanted to examine in the context of those that drink and those that do not was income since income is a large marker of inequity in adult health. The only income bracket with a higher average percent of consumers than abstainers was a high income. 

```{r}
#| label: income
#| echo: false

gen_income <- tidy_general |>
  select(indicator_behavior, wbincome2024, estimate) |>
  drop_na(wbincome2024) |>
  group_by(indicator_behavior, wbincome2024) |>
  summarize(avg_pct = mean(estimate, na.rm = TRUE), .groups = "drop") 
  
gen_income |>  
  ggplot(aes(x = indicator_behavior, y = avg_pct)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ wbincome2024, ncol = 2) +
  coord_flip() + 
  labs(
    title = "Average % of population engaging in indicator behavior by income", 
    x = "Average Percent", 
    y = "Indicator Behavior")

gen_income |>
  pivot_wider(
    names_from = "indicator_behavior", 
    values_from = "avg_pct"
  ) |>
  knitr::kable(caption = "Table 2: Average % of population engaging in indicator behavior by income")

```


## Regional Analysis

Next, I examined the different regions identified in the dataset. My main goal with this analysis was to determine which countries I might want to examine further, based on which countries seem to have higher alcohol consumption. My initial finding was that European settings have the highest average percent of consumers at 60.75%. This is also the only region where consumption is the behavior that represents the largest portion of the population. 

```{r}
#| label: regions
#| echo: false

gen_regions <- tidy_general |>
  select(indicator_behavior, whoreg6, estimate) |>
  group_by(indicator_behavior, whoreg6) |>
  summarize(avg_pct = mean(estimate, na.rm = TRUE), .groups = "drop") 

gen_regions |>
  ggplot(aes(x = indicator_behavior, y = avg_pct)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ whoreg6, ncol = 2) +
  coord_flip() + 
  labs(
    title = "Average % of population engaging in indicator behavior by region", 
    x = "Average Percent", 
    y = "Indicator Behavior")

gen_regions |>
  pivot_wider(
    names_from = "indicator_behavior", 
    values_from = "avg_pct"
  ) |>
  knitr::kable(caption = "Table 3: Average % of population engaging in indicator behavior by region")

```

## European Countries

To gain a better understanding of consumption in the European region, I gathered the average percents of each behavior in this focus by setting. I then sorted them by percent of consumers, and determined that Luxembourg has the highest average percent of consumers at 91.8%. 

```{r}
#| label: europe
#| echo: false

tidy_general |>
  select(
    indicator_behavior, 
    whoreg6, 
    setting, 
    setting_average
  ) |>
  filter(whoreg6 == "European") |>
  summarize(
    setting_average = mean(setting_average), 
    .by = c(indicator_behavior, setting)
  ) |>
  pivot_wider(
    names_from = "indicator_behavior", 
    values_from = "setting_average"
  ) |>
  arrange(desc(!!sym("% consumers, past 12 mo"))) |>
  slice_head(n = 10) |>
  knitr::kable(caption = "Table 4: Average % of population engaging in indicator behavior by European country")

```

## Summary and Next Steps

Throughout this initial EDA I have gained a better understanding of focus groups that I may want to further examine using other indicator behaviors. Specifically in these higher consuming european settings, I would like to examine the level to which drinking is occurring and if there is fatality involved. 

I think it would be beneficial, however, to find where the highest-consuming settings are within each reagion to examine through multivariate analysis if there are connections to variables like gender, age, or income. I think it will be useful to subset the tidied data into categories of behaviors a I did for this more general analysis, accounting for behaviors associated with given ages or outcomes to organize myself as I carry our these analyses. 

I would also like to make use of maps to facilitate these visualizations. I think it would be valuable to both pinpoint where certain behaviors are present and also the distribution of values throughout the settings. 
